Generalized Measurement Invariance Tests for Factor Analysis

### 2012-01-23 www/psychoco2012/invariance.tex
Minor revisions to Psychoco 2012 slides, along with TODO's for revisions that will take longer than 2min.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@88 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

% FA path diagram, types of MI
\frame{
  \frametitle{Studying Measurement Invariance}
  % path diagram illustrating parameters being the same for everyone
  \begin{center}
      \includegraphics[height=3in]{famod.pdf}
  \end{center}}

% Define as theta_i = theta_0, i=1,..,n
\frame{
  \frametitle{Studying Measurement Invariance}
  \begin{itemize}
    % \item There exist many types of measurement invariance,
    %   based on specific subsets of parameters being allowed
    %   to vary across individuals (e.g., Meredith, 1993).\\ \ \\
    \item Hypothesis of ``full'' measurement invariance:\\
      \begin{center}
          $H_0: \bm{\theta}_i = \bm{\theta}_0, i=1,\ldots,n$\\
          $H_1: \text{Not all the }\bm{\theta}_i = \bm{\theta}_0$
      \end{center}
  \end{itemize}}

@@

\frame{
  \frametitle{Proposed Tests}
  \begin{itemize}
    \item The proposed family of tests rely on first derivatives of
      the model's log-likelihood function.
  % Equation of gradient vs scores for individuals
  \end{itemize}
\begin{eqnarray*}
  \displaystyle\sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}}) =
  {\bm 0},\text{ where} \\
 \\
  \psi(\bm{x}_i, {\bm \theta}) = \frac{\partial}{\partial \bm{\theta}} \log
\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
\end{eqnarray*}}

\frame{
  \frametitle{Proposed Tests}
  \begin{itemize}
    \item We can also consider individual terms ({\em{scores}}) of the
      gradient.  These scores tell us how well a particular
      parameter describes a particular individual.  If your score is
      0 for some parameter, that parameter describes you well.  If your score
      is far from 0, that parameter describes you poorly. \\ \ \\
  \end{itemize}
  \begin{equation*}
  \psi(\bm{x}_i, {\bm \theta}) = \frac{\partial}{\partial \bm{\theta}} \log
  \text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}      
  \end{equation*}}

@@

\frame{
  \frametitle{Theorem}
 % show that cumulative scores follow a Brownian bridge under H_0
  \begin{itemize}
    \item Theorem (Zeileis \& Hornik, 2007): Under the hypothesis of
      measurement invariance, the following functional central limit
      theorem holds:
  \end{itemize}
  \begin{center}
  ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}}) 
\overset{d}{\rightarrow} {\bm W}^{0}(\cdot),$\\ \ \\
 where ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}$ is the observed
 information matrix and ${\bm W}^{0}(\cdot)$ is a
 $p$-dimensional Brownian bridge.
\end{center}}

@@

% Picture of Brownian bridge
\frame{
  \frametitle{Brownian Bridge}
  \begin{center}
      \includegraphics[height=3in]{brownian.pdf}
  \end{center}}

@@

\frame{
  \frametitle{Software}
  \begin{itemize}
    \item To carry out the tests, we use
      \begin{itemize}
        \item \texttt{lavaan} for model estimation.
        \item A combination of \texttt{lavaan} and our own code for calculating scores of fitted models.
        \item \texttt{strucchange} for carrying out the proposed tests with the scores.
      \end{itemize}
  \end{itemize}}

@@

% Insert new results
\frame{
  \frametitle{Results}
  \begin{center}
      \includegraphics[height=3in]{paper-gefp-4-1.pdf}
  \end{center}}

@@

\frame{
  \frametitle{Current Work}
  \begin{itemize}
    \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
    \item Detailed examination of test properties via simulation.\\ \ \\
    \item Extension to related psychometric issues (modification
      indices, mediation).
  \end{itemize}}

>>>>

% TODO: Add something like Fig 3 from paper to show test stat construction?

@@

% FA path diagram, types of MI
% TODO: Overlays with different parameter values on graphic
\frame{
  \frametitle{Studying Measurement Invariance}
  % path diagram illustrating parameters being the same for everyone
  \begin{center}
      \includegraphics[height=3in]{famod.pdf}
  \end{center}}

% Define as theta_i = theta_0, i=1,..,n
\frame{
  \frametitle{Studying Measurement Invariance}
  \begin{itemize}
    % \item There exist many types of measurement invariance,
    %   based on specific subsets of parameters being allowed
    %   to vary across individuals (e.g., Meredith, 1993).\\ \ \\
    \item Hypothesis of ``full'' measurement invariance ($\bm{\theta}_i$ is the full parameter vector for individual $i$):\\
      \begin{center}
          $H_0: \bm{\theta}_i = \bm{\theta}_0, i=1,\ldots,n$\\
          $H_1: \text{Not all the }\bm{\theta}_i = \bm{\theta}_0$
      \end{center}
  \end{itemize}}

@@

% TODO Consistent notation with paper
\frame{
  \frametitle{Proposed Tests}
  \begin{itemize}
    \item The proposed family of tests rely on first derivatives of
      the model's log-likelihood function.
    \item We can also consider individual terms ({\em scores}) of the gradient.  These scores tell us how well a particular parameter describes a particular individual.
  % Equation of gradient vs scores for individuals
  \end{itemize}
\begin{eqnarray*}
  \displaystyle\sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}}) =
  {\bm 0},\text{ where} \\
 \\
  \psi(\bm{x}_i, \hat{{\bm \theta}}) = \frac{\partial}{\partial \bm{\theta}} \log
\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
\end{eqnarray*}}

% \frame{
%   \frametitle{Proposed Tests}
%   \begin{itemize}
%     \item We can also consider individual terms ({\em{scores}}) of the
%       gradient.  These scores tell us how well a particular
%       parameter describes a particular individual.  If your score is
%       0 for some parameter, that parameter describes you well.  If your score
%       is far from 0, that parameter describes you poorly. \\ \ \\
%   \end{itemize}
%   \begin{equation*}
%   \psi(\bm{x}_i, {\bm \theta}) = \frac{\partial}{\partial \bm{\theta}} \log
%   \text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}      
%   \end{equation*}}

@@

\frame{
  \frametitle{Theorem}
 % show that cumulative scores follow a Brownian bridge under H_0
  \begin{itemize}
    \item Theorem (Zeileis \& Hornik, 2007): Under the hypothesis of
      measurement invariance, the following functional central limit
      theorem holds:
  \begin{equation}
  ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}}) 
\overset{d}{\rightarrow} {\bm W}^{0}(\cdot),$
  \end{equation}
 where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed
 information matrix and ${\bm W}^{0}(\cdot)$ is a
 $p$-dimensional Brownian bridge (where $p$ is the number of parameters).
  \end{itemize}
}

@@

% Picture of Brownian bridge
% \frame{
%   \frametitle{Brownian Bridge}
%   \begin{center}
%       \includegraphics[height=3in]{brownian.pdf}
%   \end{center}}

@@

@@

% TODO: Break up results over multiple slides
\frame{
  \frametitle{Results}
  \begin{center}
      \includegraphics[height=3in]{paper-gefp-4-1.pdf}
  \end{center}}

@@

% TODO: Add some detail on software
\frame{
  \frametitle{Software}
  \begin{itemize}
    \item To carry out the tests, we use
      \begin{itemize}
        \item \texttt{lavaan} for model estimation.
        \item A combination of \texttt{lavaan} and our own code for calculating scores of fitted models.
        \item \texttt{strucchange} for carrying out the proposed tests with the scores.
      \end{itemize}
  \end{itemize}}

\frame{
  \frametitle{Current Work}
  \begin{itemize}
    \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
    \item Detailed examination of test properties via simulation.\\ \ \\
    \item Extension to related psychometric issues (modification
      indices, mediation).
   % TODO: URL for working paper
  \end{itemize}}
